#Setup
```{r, warning = F, message = F}
packages <- c("tidyverse", "vegan", "leaflet", "readxl", "lme4")
lapply(packages, library, character.only = T)
theme_set(theme_classic(base_size = 24))

#Colour palette: "#663300" (brown), "#f51a0a" (red), "#ff00bf" (pink), "#2eb82e" (green), "#2763c4" (blue)
```

#Managing the mushroom observations data

```{r, warning = F, message = F}
#MushroomObserver Main File
obs <- read_tsv("observations.csv")
# Removing the useless information
obs <- obs %>% 
  subset(is_collection_location == 1) %>% # Some reported specimens were not picked at that location 
  select(name_id, when, location_id) #Taking only useful columns

#--------------------------------------------

# Reading in the location data for MushroomObserver
locCrude <- read_tsv("locations.csv")

# Including the location, name, id, and latitude and longitude
# Note the file gives north, east, south and west bounds, we take the center of this region
# Making column names suitable for the observation file
loc <- locCrude %>% 
  mutate(lat = (north + south) / 2, long = (west + east) / 2) %>% 
  select(location_id = id, lat, long, location = name)

# Using the location data to identify the locations ids in the observation file
obs <- left_join(obs, loc, by = "location_id")

#--------------------------------------------

# Reading in the identification data for MushroomObserver
speciesCrude <- read_tsv("names.csv")
furtherTaxonomy <- read_tsv("name_classifications.csv")[,c(1,7)]

# Including the name id, name, and taxonomic rankings
species <- speciesCrude[,c(1,2,7)]
species <- rename(species, name_id = id)

# Using the name data to identify the name ids in the observation file
obs <- left_join(obs, species, by = "name_id")
obs <- left_join(obs, furtherTaxonomy, by = "name_id")

colnames(obs) <- c("nameId", "when", "locationId", 
                          "lat", "lon", "location", "name", "rank", "family")
# The rank indicates what taxonomic level the observer provides (species, genus, family, etc.) - species is 4, genus is 9

```

#Getting a sense of the data with Leaflet
```{r}
#Grouping observations into more digestible (and readable/processable) chunks of time
mapObs <- obs %>% select(when, lat, lon) %>% 
  mutate(timeframe = cut(when, as.Date(c("1880-01-01", "2000-01-01", "2002-01-01", "2004-01-01",
                                 "2006-01-01", "2008-01-01", "2010-01-01", "2012-01-01",
                                 "2014-01-01", "2016-01-01", "2018-01-01", "2020-01-01", 
                                 "2022-01-01", "2023-01-01")),
                         labels = c("ancient", "2000-02", "2002-04", "2004-06", "2006-08",
                                    "2008-10", "2010-12", "2012-14", "2014-16", "2016-18",
                                    "2018-20", "2020-22", "2022-23")))
mapObs <- split(mapObs, mapObs$timeframe)

#Creating the leaflet
##To be honest, not entirely sure how all this code works - it just does
l <- leaflet() %>% addTiles() 
names(mapObs) %>% 
  walk(function(df) {
    l <<- l %>% 
      addMarkers(data = mapObs[[df]],
                 ~lon, ~lat,
                 label = ~as.character(when),
                 popup = ~as.character(when),
                 group = df,
                 clusterOptions = markerClusterOptions(removeOutsideVisibleBounds = F),
                 labelOptions = labelOptions(noHide = F, direction = "auto")
                 )
  })

#Viewing the leaflet with a legend control
l %>% addLayersControl(
  overlayGroups = names(mapObs),
  options = layersControlOptions(collapsed = FALSE)
)

```

#Climate 

##San Francisco
```{r}
# Filtering observations for only observations seen in the 
# San Francsico bay area during the climate data window
sanFranObs <- obs %>% 
  filter(lat <= 38 & lat >= 37.7 & 
           lon <= -122.1 & lon >= -122.8 &
           when >= as.Date("2007-01-01") & when < as.Date("2023-01-01"))
```

###Monthly data manipulation 

```{r}
# San Francisco monthly climate average
sanFranClimate <- read.table("MonthClim.txt")
sanFranClimate <- sanFranClimate %>% 
  filter(V2 != -9999) %>% # Removing NA values
  select(when = V1, temp = V2, precipitation = V7) %>% 
  mutate(when = my(when)) # Making the data type a date

# Grouping by month
# Converting to abundance data 
# and matching to climate data
sanFranAbundance <- sanFranObs %>%   
  group_by(when = floor_date(when, "month")) %>% 
  tally() %>% 
  left_join(sanFranClimate, by = "when") %>% 
  filter(!is.na(temp))

# Grouping by month
# Converting to data suitable for diversity calculations
sanFranDiversity <- sanFranObs %>% 
  filter(!is.na(family)) %>% # Modifiable for species (= 4), genus (= 9), family (!is.na(family))
  mutate(genus = lapply(strsplit(name, split = " "), '[[', 1)) %>% # To target for genus
  select(when, family) %>% # Can select for name, genus, or family
  group_by(when = floor_date(when, "month"), family) %>% # Can group by name, genus, or family
  tally() %>% 
  pivot_wider(names_from = family, values_from = n, 
              values_fill = 0) # Can be name, genus, or family
  
## Calculating monthly diversity and matching to climate
sanFranDiversity <- data.frame(sanFranDiversity, 
                          diversity = diversity(sanFranDiversity[,2:length(sanFranDiversity)], 
                                                index = "shannon")) %>% 
  left_join(sanFranClimate, by = "when") %>% 
  filter(!is.na(temp))
```

###Monthly Exploration

####Abundance
```{r}
# Re-making the data "messy"
## Making a continuous time scale so ggplot plots blank space where there are
## null values instead of connecting nonconsecutive points with an ugly line
timeseries <- data.frame(when = seq.Date(min(sanFranAbundance$when), max(sanFranAbundance$when), by = "month"))
abundanceTimeseries <- full_join(timeseries, sanFranAbundance, by = "when")

# Plotting monthly observations to monthly climate
abundanceTimeseries %>% 
  ggplot(aes(x = when)) + 
  geom_line(aes(y = n, col = "Observations"), size=0.75) +
  geom_line(aes(y = precipitation, col = "Precipitation"), size=0.75) +
  geom_line(aes(y = temp/0.05, col = "Temperature"), size=0.75) +
  scale_y_continuous(sec.axis = sec_axis(~.*0.05, name = "Temperature (°C)")) +
  labs(y = "Abundance (n) and Precipitation (mm)", x = element_blank()) +
  xlim(x = as.Date("2010-01-01"), as.Date("2020-12-01")) +
  scale_colour_manual(values = c("#2eb82e", "#2763c4", "#f51a0a"),
                    name = element_blank(),
                    breaks = c("Observations", "Precipitation", "Temperature"),
                    labels = c("Observations", "Precipitation", "Temperature"))


# Examining the correlations between observations and temperature/precipitation
sanFranAbundance %>% 
  ggplot(aes(x = temp, y = n)) +
  geom_point(col = "#f51a0a") + 
  geom_smooth(method = "glm", method.args = list(family = "poisson"), col = "red") +
  labs(x = "Temperature (°C)", y = "Abundance (n)")

sanFranAbundance %>% 
  ggplot(aes(x = precipitation, y = n)) +
  geom_point(col = "#2763c4") + 
  geom_smooth(method = "glm", method.args = list(family = "poisson"), col = "blue") +
  labs(x = "Precipitation (mm)", y = "Abundance (n)")

sanFranAbundance %>% 
  ggplot(aes(x = temp, y = precipitation, colour = n)) +
  geom_point() +
  scale_colour_gradient(low = "#2eb82e", high = "#48742F", name = "Abundance (n)") +
  labs(x = "Temperature (°C)", y = "Precipitation (mm)")
```

####Diversity

```{r}
# Re-making the data "messy"
## Making a continuous time scale so ggplot plots blank space where there are
## null values instead of connecting nonconsecutive points with an ugly line
timeseries <- data.frame(when = seq.Date(min(sanFranDiversity$when), max(sanFranDiversity$when), by = "month"))
diversityTimeseries <- full_join(timeseries, sanFranDiversity, by = "when")

# Diversity with climate overlayed - diversity values scaled up for clarity
diversityTimeseries %>% 
  ggplot(aes(x = when)) + 
  geom_line(aes(y = diversity*100, col = "Diversity"), size=0.75) +
  geom_line(aes(y = precipitation, 
                col = "Precipitation"), size=0.75) +
  geom_line(aes(y = temp/0.05, 
                col = "Temperature"), size=0.75) +
  scale_y_continuous(sec.axis = sec_axis(~.*0.05, name = "Temperature (°C)")) +
  labs(y = "Diversity (x100) and Precipitation (mm)", x = element_blank()) +
  xlim(x = as.Date("2010-01-01"), as.Date("2020-12-01")) +
  scale_colour_manual(values = c("#2eb82e", "#2763c4", "#f51a0a"),
                    name = element_blank(),
                    breaks = c("Diversity", "Precipitation", "Temperature"),
                    labels = c("Diversity", "Precipitation", "Temperature"))

# Examining the correlations between observations and temperature/precipitation
sanFranDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = temp, y = diversity)) +
  geom_point(col = "#f51a0a") + 
  geom_smooth(method = "glm", method.args = list(family = "Gamma"), col = "red") +
  labs(x = "Temperature (°C)", y = "Diversity")

sanFranDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = precipitation, y = diversity)) +
  geom_point(col = "#2763c4") + 
  geom_smooth(method = "glm", method.args = list(family = "Gamma"), col = "blue") +
  labs(x = "Precipitation (mm)", y = "Diversity") +
  ylim(0, 4)

sanFranDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = temp, y = precipitation, colour = diversity)) +
  geom_point() +
  scale_colour_gradient(low = "#2eb82e", high = "#48742F", name = "Diversity") +
  labs(x = "Temperature (°C)", y = "Precipitation (mm)")
```



##Ontario
```{r}
# Filtering observations for only observations seen in Ontario 
# during the climate data window
ontarioObs <- obs %>% 
  filter(lat <= 43.85 & lat >= 42.54 & 
           lon <= -79 & lon >= -82.4 &
           when >= as.Date("2010-01-01") & when < as.Date("2023-01-01"))
```

###Monthly data manipulation 

```{r}
# Ontario monthly climate average
ontarioClimate <- read_csv("ontarioClimate.csv")[,c(19,6,32)] %>% 
  group_by(when = floor_date(ymd(LOCAL_DATE), "month")) %>% 
  subset(!is.na(MEAN_TEMPERATURE) & !is.na(TOTAL_PRECIPITATION)) %>% 
  summarize(temp = mean(MEAN_TEMPERATURE), 
            precipitation = sum(TOTAL_PRECIPITATION))

# Converting to abundance data and matching to climate data
ontarioAbundance <- ontarioObs %>%   
  group_by(when = floor_date(when, "month")) %>% 
  tally() %>% 
  left_join(ontarioClimate, by = "when") %>% 
  filter(!is.na(temp))

# Converting to data suitable for the diversity function
ontarioDiversity <- ontarioObs %>% 
  filter(!is.na(family)) %>% # Modifiable for species (= 4), genus (= 9), family (!is.na(family))
  mutate(genus = lapply(strsplit(name, split = " "), '[[', 1)) %>% # To target for genus
  select(when, family) %>% # Can be name, genus, or family
  group_by(when = floor_date(when, "month"), family) %>% # Can be name, genus, or family
  tally() %>% 
  pivot_wider(names_from = family, values_from = n, 
              values_fill = 0) # Can be name, genus, or family
  
## Calculating diversity and matching to climate
ontarioDiversity <- data.frame(ontarioDiversity, 
                          diversity = diversity(ontarioDiversity[,2:length(ontarioDiversity)], 
                                                index = "shannon")) %>% 
  left_join(ontarioClimate, by = "when") %>% 
  filter(!is.na(temp))
```

###Monthly Exploration

####Abundance
```{r}
# Re-making the data "messy"
## Making a continuous time scale so ggplot plots blank space where there are
## null values instead of connecting nonconsecutive points with an ugly line
timeseries <- data.frame(when = seq.Date(min(ontarioAbundance$when), max(ontarioAbundance$when), by = "month"))
abundanceTimeseries <- full_join(timeseries, ontarioAbundance, by = "when")

# Plotting monthly observations to monthly climate
abundanceTimeseries %>% 
  ggplot(aes(x = when)) + 
  geom_line(aes(y = n, col = "Observations"), size=0.75) +
  geom_line(aes(y = precipitation, col = "Precipitation"), size=0.75) +
  geom_line(aes(y = temp/0.05, col = "Temperature"), size=0.75) +
  scale_y_continuous(sec.axis = sec_axis(~.*0.05, name = "Temperature (°C)")) +
  labs(y = "Abundance (n) and Precipitation (mm)", x = element_blank()) +
  scale_colour_manual(values = c("#2eb82e", "#2763c4", "#f51a0a"),
                    name = element_blank(),
                    breaks = c("Observations", "Precipitation", "Temperature"),
                    labels = c("Observations", "Precipitation", "Temperature"))

# Examining the correlations between observations and temperature/precipitation
ontarioAbundance %>% 
  ggplot(aes(x = temp, y = n)) +
  geom_point(col = "#f51a0a") + 
  geom_smooth(method = "glm", method.args = list(family = "poisson"), col = "red") +
  labs(x = "Temperature (°C)", y = "Abundance (n)")

ontarioAbundance %>% 
  ggplot(aes(x = precipitation, y = n)) +
  geom_point(col = "#2763c4") + 
  geom_smooth(method = "glm", method.args = list(family = "poisson"), col = "blue") +
  labs(x = "Precipitation (mm)", y = "Abundance (n)")

ontarioAbundance %>% 
  ggplot(aes(x = temp, y = precipitation, colour = n)) +
  geom_point() +
  scale_colour_gradient(low = "#2eb82e", high = "#48742F", name = "Abundance (n)") +
  labs(x = "Temperature (°C)", y = "Precipitation (mm)")
```

####Diversity

```{r}
# Re-making the data "messy"
## Making a continuous time scale so ggplot plots blank space where there are
## null values instead of connecting nonconsecutive points with an ugly line
timeseries <- data.frame(when = seq.Date(min(ontarioDiversity$when), max(ontarioDiversity$when), by = "month"))
diversityTimeseries <- full_join(timeseries, ontarioDiversity, by = "when")

# Diversity with climate overlayed - diversity values scaled up for clarity
diversityTimeseries %>% 
  ggplot(aes(x = when)) + 
  geom_line(aes(y = diversity*100, col = "Diversity"), size=0.75) +
  geom_line(aes(y = precipitation, 
                col = "Precipitation"), size=0.75) +
  geom_line(aes(y = temp/0.05, 
                col = "Temperature"), size=0.75) +
  scale_y_continuous(sec.axis = sec_axis(~.*0.05, name = "Temperature (°C)")) +
  labs(y = "Diversity (x100) and Precipitation (mm)", x = element_blank()) +
  scale_colour_manual(values = c("#2eb82e", "#2763c4", "#f51a0a"),
                    name = element_blank(),
                    breaks = c("Diversity", "Precipitation", "Temperature"),
                    labels = c("Diversity", "Precipitation", "Temperature"))

# Examining the correlations between observations and temperature/precipitation
ontarioDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = temp, y = diversity)) +
  geom_point(col = "#f51a0a") + 
  geom_smooth(method = "glm", method.args = list(family = "Gamma"), col = "red") +
  labs(x = "Temperature (°C)", y = "Diversity")

ontarioDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = precipitation, y = diversity)) +
  geom_point(col = "#2763c4") + 
  geom_smooth(method = "glm", method.args = list(family = "Gamma"), col = "blue") +
  labs(x = "Precipitation (mm)", y = "Diversity")

ontarioDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = temp, y = precipitation, colour = diversity)) +
  geom_point() +
  scale_colour_gradient(low = "#2eb82e", high = "#48742F", name = "Diversity") +
  labs(x = "Temperature (°C)", y = "Precipitation (mm)")
```

##Colombia
```{r}
# Filtering observations for only observations seen in Colombia 
# during the climate data window
colombiaObs <- obs %>% 
  filter(lat <= 12.4 & lat >= 1.2 & 
           lon <= -70 & lon >= -79 &
           when >= as.Date("2010-01-01") & when < as.Date("2023-01-01"))
```

###Monthly data manipulation 

```{r} 
# Colombia monthly climate average - data is contained within two separate files
## Temperature
colombiaTemp <- read.table("colombiaTemp.per", skip = 3, header = T) %>% 
  select(-c(MAM, JJA, SON, DJF, ANN)) %>%
  rename("01" = "JAN", "02" = "FEB", "03" = "MAR", "04" = "APR", 
         "05" = "MAY", "06" = "JUN", "07" = "JUL", "08" = "AUG", 
         "09" = "SEP", "10" = "OCT", "11" = "NOV", "12" = "DEC") %>% 
  pivot_longer(c(2:13), names_to = "Month", values_to = "temp") %>% 
  mutate(date = ym(paste(YEAR, Month, sep = "-"))) %>% 
  filter(date >= as.Date("2010-01-01"))

## Precipitation
### Merging the two datasets together into one climate dataframe
colombiaClimate <- read.table("colombiaPrecip.per", skip = 3, header = T) %>% 
  select(-c(MAM, JJA, SON, DJF, ANN)) %>%
  rename("01" = "JAN", "02" = "FEB", "03" = "MAR", "04" = "APR", 
         "05" = "MAY", "06" = "JUN", "07" = "JUL", "08" = "AUG", 
         "09" = "SEP", "10" = "OCT", "11" = "NOV", "12" = "DEC") %>% 
  pivot_longer(c(2:13), names_to = "Month", values_to = "precipitation") %>% 
  mutate(when = ym(paste(YEAR, Month, sep = "-"))) %>% 
  filter(when >= as.Date("2010-01-01")) %>% 
  select(when, precipitation) %>% 
  data.frame(colombiaTemp[3])

# Converting to abundance data and matching to climate data
colombiaAbundance <- colombiaObs %>%   
  group_by(when = floor_date(when, "month")) %>% 
  tally() %>% 
  left_join(colombiaClimate, by = "when") %>% 
  filter(!is.na(temp))

# Converting to data suitable for diversity calculation
colombiaDiversity <- colombiaObs %>% 
  filter(!is.na(family)) %>% # Modifiable for species (= 4), genus (= 9), family (!is.na(family))
  mutate(genus = lapply(strsplit(name, split = " "), '[[', 1)) %>% # To target for genus
  select(when, family) %>% # Can be name, genus, or family
  group_by(when = floor_date(when, "month"), family) %>% # Can be name, genus, or family
  tally() %>% 
  pivot_wider(names_from = family, values_from = n, 
              values_fill = 0) # Can be name, genus, or family
  
## Calculating diversity and matching to climate
colombiaDiversity <- data.frame(colombiaDiversity, 
                          diversity = diversity(colombiaDiversity[,2:length(colombiaDiversity)], 
                                                index = "shannon")) %>% 
  left_join(colombiaClimate, by = "when") %>% 
  filter(!is.na(temp))
```

###Monthly Exploration

####Abundance
```{r}
# Re-making the data "messy"
## Making a continuous time scale so ggplot plots blank space where there are
## null values instead of connecting nonconsecutive points with an ugly line
timeseries <- data.frame(when = seq.Date(min(colombiaAbundance$when), max(colombiaAbundance$when), by = "month"))
abundanceTimeseries <- full_join(timeseries, colombiaAbundance, by = "when")

# Plotting monthly observations to monthly climate
abundanceTimeseries %>% 
  ggplot(aes(x = when)) + 
  geom_line(aes(y = n, col = "Observations"), size=0.75) +
  geom_line(aes(y = precipitation, col = "Precipitation"), size=0.75) +
  geom_line(aes(y = temp/0.05, col = "Temperature"), size=0.75) +
  scale_y_continuous(sec.axis = sec_axis(~.*0.05, name = "Temperature (°C)")) +
  xlim(x = as.Date("2013-01-01"), as.Date("2022-12-01")) +
  labs(y = "Abundance (n) and Precipitation (mm)", x = element_blank()) +
  scale_colour_manual(values = c("#2eb82e", "#2763c4", "#f51a0a"),
                    name = element_blank(),
                    breaks = c("Observations", "Precipitation", "Temperature"),
                    labels = c("Observations", "Precipitation", "Temperature"))

# Examining the correlations between observations and temperature/precipitation
colombiaAbundance %>% 
  ggplot(aes(x = temp, y = n)) +
  geom_point(col = "#f51a0a") + 
  geom_smooth(method = "glm", method.args = list(family = "poisson"), col = "red") +
  labs(x = "Temperature (°C)", y = "Abundance (n)")

colombiaAbundance %>% 
  ggplot(aes(x = precipitation, y = n)) +
  geom_point(col = "#2763c4") + 
  geom_smooth(method = "glm", method.args = list(family = "poisson"), col = "blue") +
  labs(x = "Precipitation (mm)", y = "Abundance (n)")

colombiaAbundance %>% 
  ggplot(aes(x = temp, y = precipitation, colour = n)) +
  geom_point() +
  scale_colour_gradient(low = "#2eb82e", high = "#48742F", name = "Abundance (n)") +
  labs(x = "Temperature (°C)", y = "Precipitation (mm)")
```

####Diversity

```{r}
# Re-making the data "messy"
## Making a continuous time scale so ggplot plots blank space where there are
## null values instead of connecting nonconsecutive points with an ugly line
timeseries <- data.frame(when = seq.Date(min(colombiaDiversity$when), max(colombiaDiversity$when), by = "month"))
diversityTimeseries <- full_join(timeseries, colombiaDiversity, by = "when")

# Diversity with climate overlayed - diversity values scaled up for clarity
diversityTimeseries %>% 
  ggplot(aes(x = when)) + 
  geom_line(aes(y = diversity*50, col = "Diversity"), size=0.75) +
  geom_line(aes(y = precipitation, 
                col = "Precipitation"), size=0.75) +
  geom_line(aes(y = temp/0.075, 
                col = "Temperature"), size=0.75) +
  scale_y_continuous(sec.axis = sec_axis(~.*0.075, name = "Temperature (°C)")) +
  xlim(x = as.Date("2013-01-01"), as.Date("2022-12-01")) +
  labs(y = "Diversity (x50) and Precipitation (mm)", x = element_blank()) +
  scale_colour_manual(values = c("#2eb82e", "#2763c4", "#f51a0a"),
                    name = element_blank(),
                    breaks = c("Diversity", "Precipitation", "Temperature"),
                    labels = c("Diversity", "Precipitation", "Temperature"))

# Examining the correlations between observations and temperature/precipitation
colombiaDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = temp, y = diversity)) +
  geom_point(col = "#f51a0a") + 
  geom_smooth(method = "glm", method.args = list(family = "Gamma"), col = "red") +
  labs(x = "Temperature (°C)", y = "Diversity")

colombiaDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = precipitation, y = diversity)) +
  geom_point(col = "#2763c4") + 
  geom_smooth(method = "glm", method.args = list(family = "Gamma"), col = "blue") +
  labs(x = "Precipitation (mm)", y = "Diversity")

colombiaDiversity %>% subset(diversity > 0) %>% 
  ggplot(aes(x = temp, y = precipitation, colour = diversity)) +
  geom_point() +
  scale_colour_gradient(low = "#2eb82e", high = "#48742F", name = "Diversity") +
  labs(x = "Temperature (°C)", y = "Precipitation (mm)")
```

##Likelihood :) to determine diversity distribution
```{r}
#Finding the parameters for a set location (Modify for sanFran, ontario, or colombia)
##sanFranDiversity(gamma(~4.9, ~2.2), invgauss(~2.2, ~7.5), exp(~0.45)), 
##ontarioDiversity(gamma(~9.8,	~3.8), invgauss(~2.6,	~17.1), exp(~0.39))
##colombiaDiversity(gamma(~7.8, ~2.9), invgauss(~2.7, ~13.5), exp(~0.37))
div <- sanFranDiversity %>% subset(diversity > 0) %>% select(diversity)

#Gamma
##Gamma Probability Density Function
LLGamma <- function(diversity, alpha, beta){
  probGamma <- ((diversity^(alpha - 1)) * exp(-beta*diversity) * (beta^alpha)) / gamma(alpha)
  return(sum(log(probGamma)))
}

##Finding the log likelihoods
LL <- c()
gammaParams <- expand.grid(alpha = seq(0.1, 15, 0.1), beta = seq(0.1, 15, 0.1))
for (i in 1:nrow(gammaParams)){
  LL[i] <- LLGamma(div, gammaParams[i, 1], gammaParams[i, 2])
}
gammaTest <- cbind(LL, gammaParams)



#Inverse-Gaussian
##Inverse-Gaussian Probability Density Function
LLInvgauss <- function(diversity, mu, lambda) {
  probInvgauss <- sqrt(lambda/(2*pi*diversity^3)) * exp((-lambda*(diversity - mu)^2) / (2*(mu^2)*diversity))
  return(sum(log(probInvgauss)))
}


##Finding the log likelihoods
LL <- c()
invgaussParams <- expand.grid(mu = seq(0.1, 15, 0.1), lambda = seq(0.1, 20, 0.1))
for (i in 1:nrow(invgaussParams)) {
  LL[i] <- LLInvgauss(div, invgaussParams[i,1],invgaussParams[i,2])
}
invgaussTest <- cbind(LL, invgaussParams)



#Exponential
##Exponential Probability Density Function
LLExp <- function(diversity, lambda) {
  probExp <- lambda*exp(-lambda*diversity)
  return(sum(log(probExp)))
}

##Finding the log likelihoods
LL <- c()
expParams <- seq(0, 10, 0.01)
for (i in 1:length(expParams)) {
  LL[i] <- LLExp(div, expParams[i])
}
expTest <- data.frame(LL, lambda = expParams)


# Finding MLEs
MLEGamma <- subset(gammaTest, LL == max(LL)); MLEGamma
MLEInvgauss <- subset(invgaussTest, LL == max(LL)); MLEInvgauss
MLEExp <- subset(expTest, LL == max(LL) & is.finite(LL)); MLEExp
# AIC to find best distribution
AIC <- data.frame(4-2*MLEGamma[1], 4-2*MLEInvgauss[1], 2-2*MLEExp[1])
colnames(AIC) <- c("gamma", "invgauss", "exp");AIC

##Graphing
###Calculating the lines
distributions <- data.frame(cbind(div, y = 0), 
           Gamma = ((div^(MLEGamma$alpha - 1)) * exp(-MLEGamma$beta*div) *  
                      (MLEGamma$beta^MLEGamma$alpha)) / gamma(MLEGamma$alpha),
           InverseGaussian = sqrt(MLEInvgauss$lambda/(2*pi*div^3)) * 
                  exp((-MLEInvgauss$lambda*(div - MLEInvgauss$mu)^2) / 
                        (2*(MLEInvgauss$mu^2)*div)),
           Exponential = MLEExp$lambda * exp(-MLEExp$lambda*div)) 
colnames(distributions)[3:5] <- c("Gamma", "InverseGaussian", "Exponential")
distributions <- pivot_longer(distributions, cols = c(3:5), names_to = "distribution")

distributions %>%
  ggplot(aes(x = diversity, y = value, colour = distribution)) +
  geom_line(linewidth = 1) +
  geom_point(aes(x = diversity, y = y), col = "black") +
  labs(x = "Diversity") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank()) +
  scale_colour_manual(values = c("blue", "red", "orange"),
                    name = "Distibution Type",
                    breaks = c("Gamma", "InverseGaussian", "Exponential"),
                    labels = c("Gamma (AIC = 384)",
                               "Inverse Gaussian (AIC = 411)",
                               "Exponential (AIC = 506)"))

```



##Models
```{r}
#Finding the best model
##Abundance model (Poisson dist.)
##sanFranAbundance, ontarioAbundance, or colombiaAbundance
modelTest <- sanFranAbundance 
AIC(glm(n ~ temp, family = poisson, data = modelTest),
    glm(n ~ precipitation, family = poisson, data = modelTest),
    glm(n ~ precipitation + temp, family = poisson, data = modelTest),
    glm(n ~ precipitation * temp, family = poisson, data = modelTest))
##Diversity model (Gamma dist.)
##sanFranDiversity, ontarioDiversity, or colombiaDiversity
divModelTest <- ontarioDiversity %>% subset(diversity > 0) 
AIC(glm(diversity ~ temp, family = Gamma, data = divModelTest),
    glm(diversity ~ precipitation, family = Gamma, data = divModelTest),
    glm(diversity ~ precipitation + temp, family = Gamma, data = divModelTest),
    glm(diversity ~ precipitation * temp, family = Gamma, data = divModelTest))

#Selected models
summary(glm(n ~ precipitation * temp,
    family = poisson, data = sanFranAbundance))
summary(glm(n ~ precipitation + temp,  # equiv AIC in interaction and separate effect models - 
    family = poisson, data = ontarioAbundance)) #interaction non-significant
summary(glm(n ~ precipitation * temp,
    family = poisson, data = colombiaAbundance))
print("-----------------------------------------------")
summary(glm(diversity ~ precipitation * temp, 
    family = Gamma, data = subset(sanFranDiversity, diversity > 0)))
summary(glm(diversity ~ temp, #Only the precipitation model is higher in AIC, only temp significant
    family = Gamma, data = subset(ontarioDiversity, diversity > 0)))
summary(glm(diversity ~ precipitation + temp, # all equiv in AIC, only + model is significant
    family = Gamma, data = subset(colombiaDiversity, diversity > 0)))

```


#Wildfires

## Fire Mapping
```{r}
fire <- read.csv("mapdataall.csv")
# Removing useless columns & formatting date to relevant information.
fire <- fire %>% 
  select(incident_name, incident_date_created, incident_county, incident_acres_burned) %>% 
  filter(incident_county != "") %>% 
  mutate(incident_date_created=ymd_hms(incident_date_created)) %>% 
  mutate(incident_date_created=format(incident_date_created, "%Y-%m-%d"))

# Taking a quick look
ggplot(fire, aes(x=incident_county, y=incident_acres_burned)) +
  geom_bar(stat="sum") +
  labs(title="Acres burned in California Counties",
       x="County",
       y="Total Acres Burned") +
  theme(axis.text.x = element_text(angle = 90, size = 6))


# Selecting relevant columns, reformatting strings to standard, relating it to
# mushroom dataset.
caCountiesXlsx <- read_excel("us-county-boundaries.xlsx")
write.csv(caCountiesXlsx, file="ca-county-coords.csv", row.names=FALSE)
caCounties <- read.csv("ca-county-coords.csv")
caCounties <- caCounties %>% 
  select(NAMELSAD, INTPTLAT, INTPTLON) %>% 
  mutate(NAMELSAD=sub(" County", "", NAMELSAD))


# merged fire and ca_counties to correlate longitude and latitude based on county
mergedFire <- fire %>% 
  left_join(caCounties, by=c("incident_county"="NAMELSAD")) %>% 
  filter(!is.na(INTPTLAT))

colnames(mergedFire) <- c("incidentName", "incidentDateCreated", "incidentCounty", 
                          "incidentAcresBurned", "lat", "lon"); mergedFire

# visual map where fires occurred based of longitudes and latitudes in the 
# merged_fire dataset.
map <- leaflet(mergedFire) %>% 
  addTiles() %>% 
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    weight = 1,
    radius = ~sqrt(incidentAcresBurned)*0.02,
    color = "red",
    fillOpacity = 0.4,
    popup = ~paste("Acres Burned: ", incidentAcresBurned)
  ) %>%
  addLegend("bottomright", colors="red", labels="Acres Burned"); map

```

## Exploring effect

```{r}
# Looking for the biggest fire
mergedFire %>% 
  arrange(desc(incidentAcresBurned)) 

# Looking through Mushroom Observer 
obs %>% 
  group_by(location) %>% 
  tally() %>% 
  arrange(desc(n))

# Graphing the trend in Observations before and after fire
## Bar
obs %>% 
  filter(location == "Yosemite National Park, Mariposa Co., Mono Co., Madera Co., and Tuolumne Co., California, USA") %>% 
  mutate(before_fire = when < "2013-08-17") %>% 
  group_by(when, before_fire) %>% 
  filter(when > "2011-08-17" & when < "2015-08-17") %>% 
  tally() %>% 
  ggplot(aes(x = when, fill = before_fire)) + geom_histogram() + geom_vline(xintercept = as.numeric(as.Date("2013-08-17"))) + labs(title = "Mushroom abundance before and after Rim Fire", x = "Dates", y = "Mushroom Count")

## Line
obs %>% 
  filter(location == "Yosemite National Park, Mariposa Co., Mono Co., Madera Co., and Tuolumne Co., California, USA") %>% 
  mutate(before_fire = when < "2013-08-17") %>% 
  group_by(when, before_fire) %>% 
  filter(when > "2011-08-17" & when < "2015-08-17") %>% 
  tally() %>% 
  ggplot(aes(x = when, y = n)) + geom_line() + geom_vline(xintercept = as.numeric(as.Date("2013-08-17"))) + labs(title = "Mushroom abundance before and after Rim Fire", x = "Dates", y = "Mushroom Count")
```

## Analysis/Search function

```{r}
# search function for QOL and ease of access to data set
# error threshold for latitude and longitude set to a ~60-70 mile radius
# for viable search results.

fireSearch <- function(dataset, search_value=NULL, lat=NULL, lon=NULL, 
                       error_threshold=1){
  if (is.null(search_value) && (is.null(lat) || is.null(lon))) {
    stop("Provide either 'search_value' or 'lat' and 'lon'")
  }
  
  if (!is.null(search_value)) {
    search_value <- tolower(search_value)
    result <- dataset[tolower(dataset$incidentCounty) == search_value, ]
  }
  
  else {
    # Latitude and longitude search threshold to find close proximity fires.
    result <- dataset[
      abs(dataset$lat - lat) <= error_threshold & 
      abs(dataset$lon - lon) <= error_threshold,
    ]
  }

  return(result)
}
mergedFire

# example cases
head(fireSearch(mergedFire, "Shasta"))
head(fireSearch(mergedFire, "shasta"))
head(fireSearch(mergedFire, "San Diego"))
head(fireSearch(mergedFire, lat=40, lon=-122))
```

## Fire statistical testing 

```{r}
# Codes to total up the area burned by fire and the number of mushroom observations

fireData <-
  mergedFire %>% 
  group_by(incidentCounty, year = year(incidentDateCreated)) %>% 
  summarise(totalAreaBurnt = sum(incidentAcresBurned))

observationGenerator <- function(county){

  output <-
  obs %>% 
  filter(str_detect(location, county)) %>% 
  group_by(year = year(when)) %>% 
  count() %>% 
  mutate(incidentCounty = county)
  
print(output)
}

# Codes for merging the fire and mushroom data/ Codes for randomly selecting 15 counties

merger <- function(county){
  
  merged <-
  merge(fireData, county, by = c("year","incidentCounty"))
  
print(merged)
}

countyData <-
as.data.frame(
  unique(fireData$incidentCounty))

countyData[sample(nrow(countyData), 15), ]

regressionData <-
rbind(merger(observationGenerator("Riverside")),merger(observationGenerator("San Diego")),merger(observationGenerator("Kern")),merger(observationGenerator("San Luis Obispo")),merger(observationGenerator("Shasta")),merger(observationGenerator("Butte")),merger(observationGenerator("Siskiyou")),merger(observationGenerator("Fresno")),merger(observationGenerator("San Bernardino")),merger(observationGenerator("Los Angeles")),merger(observationGenerator("Humboldt")), merger(observationGenerator("Marin")), merger(observationGenerator("Mendocino")), merger(observationGenerator("Santa Clara")), merger(observationGenerator("Sonoma")))

# For changing the unit from acres to square km
regressionData <- regressionData %>% 
  mutate(squareKmBurnt = totalAreaBurnt/247.1)

# Codes for statistical tests and graphing

mixedModel <- lmer(n^(1/4) ~ squareKmBurnt  + (1|incidentCounty), data = regressionData, REML = FALSE)
summary(mixedModel)

regressionData <- regressionData %>% 
  na.omit() %>% 
  mutate(fourthRoot = n^(1/4)) %>% 
  mutate(fit_InterceptOnly = predict(mixedModel))

ggplot(regressionData, aes(x = squareKmBurnt, y = fourthRoot, color = incidentCounty)) + geom_line(aes(y=fit_InterceptOnly)) + geom_point() + geom_abline(aes(intercept=2.812, slope=7.655e-05)) + labs(x = "Area Burned by Fire (Square Kilometre)", y = "Number of Mushroom Observations^(1/4)") + theme_classic() + theme(legend.key.size = unit(4, "mm"))
```

